---
title: "PCAngsd"
author: "J. Ignacio Lucas Lledó"
date: "8/12/2021"
output: html_document
bibliography: references.bib
---

Here I will use PCAngsd [@Meisner2018] and possibly also NGSadmix [@Skotte2013]
to determine population structure. Both require genotype likelihoods for all
individuals. I can generate them directly from BAM files with ANGSD.

Because we already know that *Hemiechinus* and *Atelerix* are different genera,
I'd rather exclude them from the analysis, so that we focus on the *Erinaceus*
subjects.

```{bash fromBAM}
if [ ! -e popfile.txt ]; then
   if [ ! -e make_popfile.sh ]; then cp ../2021-12-06/make_popfile.sh ./; fi
   ./make_popfile.sh
fi
if [ ! -e lowest15.txt ]; then
   cp ../2021-12-01/lowest15.txt ./
fi
if [ ! -e bestErinaceus.txt ]; then
   grep -v -f lowest15.txt popfile.txt | \
   grep -v -E "(atelerix|hemiechinus)" | cut -f 1 > bestErinaceus.txt
fi

# I will use only forward reads, to save time and memory
BAMDIR='../2021-11-22'
PCANGSD=~/bin/pcangsd/pcangsd.py

if [ ! -e bam.filelist ]; then
   ls -1 $BAMDIR/*.bam | grep -f bestErinaceus.txt > bam.filelist
fi

if [ ! -e all.beagle.gz ]; then
   angsd -GL 1 -minInd 57 -doGlf 2 -minMapQ 30 -minQ 20 \
         -doMaf 1 -ref ../../data/reference.fa -SNP_pval 1.0e-6 -rmTriallelic 0.05 \
         -doMajorMinor 1 \
         -out all -nThreads 40 -bam bam.filelist 1> all.beagle.log 2> all.beagle.err
fi

if [ ! -e all.cov ]; then
   python $PCANGSD -beagle all.beagle.gz \
                   -minMaf 0.05 \
                   -threads 40 \
                   -out all 1> all.cov.log 2> all.cov.err
fi

# Using K=5, expecting: 2 europaeus, 1 roumanicus, 1 concolor, 1 amurensis.
if [ ! -e all.qopt ]; then
   NGSadmix -likes all.beagle.gz \
            -K 5 \
            -minMaf 0.05 \
            -outfiles all \
            -P 40 1> all.ngsadmix.log 2> all.ngsadmix.err
fi

```

I assume that the output of both PCAngsd and NgsAdmix keep the order of the
samples as in the beagle-formated input file, which in turn should be the order
in which BAM files were specified.

```{r PCA}
pop <- read.table('popfile.txt', col.names = c('ind','pop'),
                  colClasses = c('character', 'factor'), row.names = 1)
bamlist <- scan('bam.filelist', what = 'character')
sampleList <- sapply(bamlist, function(x) {
  sub('.bam', '', unlist(strsplit(x, '/', fixed = TRUE))[3], fixed = TRUE) 
}, USE.NAMES = FALSE)

Erinaceus <- pop[sampleList, , drop = FALSE]
Erinaceus$pop <- droplevels(Erinaceus$pop)
byPop <- order(Erinaceus$pop)

COV <- as.matrix(read.table('all.cov'))
E   <- eigen(COV)
barplot(E$values[1:5] / sum(E$values), ylab = 'Explained variance', xlab = 'PC')
plot(E$vectors[, 1:2], col = Erinaceus$pop, xlab = 'PC 1', ylab = 'PC 2')
legend('bottomright', fill = 1:6, levels(Erinaceus$pop))
```

```{r NgsAdmix, fig.width=10}
library('RColorBrewer')
Q <- read.table('all.qopt')
#row.names(Q) <- sampleList
Q2 <- round(Q,3)
row.names(Q) <- paste(sampleList, ' (', Erinaceus$pop, '?)', sep = '')
# manually came up with this ordering...
z <- order(rowSums(Q2[,c(5,1,3)]), Q2[,2], rowSums(Q2[,c(5,1)]), Q2[,5], Q2[,3], decreasing=TRUE)
barplot(t(Q[z,c(5,1,3,2,4)]), col = brewer.pal(5, 'Set3'), border = NA,
        xlab = '', ylab = 'Admixture proportions for K=5', las=2, cex.names=0.5)
```

Well, I need to check the effect of a more aggressive filtering of sites. This does not
convince me yet.


## References