---
title: "Diversity and divergence in loci"
author: "J. Ignacio Lucas Lledó"
date: "15/6/2020"
output: html_document
---

```{r setup, include=TRUE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(pegas)
library(knitrProgressBar)
pop <- read.table('../../data/populations.txt', col.names=c('ind','popCode'), row.names=1)
pop$pop <- factor('roumanicus',
                  levels=c('roumanicus','europaeus','concolor','hybrid','Hemiechinus','Atelerix'))
pop[pop$popCode == 2, 'pop'] <- 'europaeus'
pop[pop$popCode == 3, 'pop'] <- 'concolor'
pop[pop$popCode == 4, 'pop'] <- 'hybrid'
pop[pop$popCode == 5, 'pop'] <- 'Hemiechinus'
pop[pop$popCode == 6, 'pop'] <- 'Atelerix'

pop2 <- pop
rownames(pop2) <- paste(rownames(pop), '0', sep='_')
pop3 <- pop
rownames(pop3) <- paste(rownames(pop), '1', sep='_')
pop2 <- rbind(pop2, pop3)
pop2 <- pop2[,2,drop=FALSE]
rm(pop3)
```

I need to extract from every alignment: the number of segregating sites
per population, the average number of differences between pairs of populations,
and the alignment length. I consider only *E. europaeus*, *E. roumanicus*, and
*E. concolor*. The strategy is the following: remove positions with gaps from the
alignment, deal with unknown alleles (Ns), remove invariable sites, calculate
haplotype frequencies in populations, calculate Hamming distances among haplotypes,
and finally get the numbers.

I will impute missing alleles with the major one if the major frequency is large
and the site does not have more than 2 missing alleles. Otherwise, the site
will be removed.

The distance between populations is the average or expected number of nucleotide
differences between two randomly picked sequences from the two populations. I use
a *for* loop (sorry) to iterate across haplotype pairs. The formula could be this:

$D_{xy} = \frac{1}{n_x n_y} \cdot \sum_{k=1}^H \sum_{l \neq k} (F_{xk}F_{yl} + F_{xl}F_{yk}) \cdot d_{kl}$

where $n_i$ is the number sequences in population $i$; $F_{ij}$ is the absolute
frequency of haplotype $j$ in population $i$, and $d_{kl}$ is the Hamming distance
between haplotypes $k$ and $l$. 

```{r functions}
diverdiver <- function(x, pop) {
  stopifnot(class(x) == 'DNAbin',
            sum(labels(x) %in% rownames(pop)) > 5)

  # Removes gap columns, unless gaps are less frequent than threshold
  x <- del.colgapsonly(x, threshold=0.0001)
  
  xChar <- as.character(x)
  xChar[xChar %in% c('n','N','-')] <- NA
  
  # Keeps only individuals in pop with less than 0.1 missing alleles
  filterInd <- apply(xChar, 1, function(x) sum(is.na(x))) / dim(xChar)[1] < 0.1 &
    rownames(xChar) %in% rownames(pop)
  stopifnot(sum(filterInd) > 5)
  filterPos <- apply(xChar, 2, function(x) sum(is.na(x))) / dim(xChar)[2] < 0.1
  stopifnot(sum(filterPos) > 0)
  xChar <- xChar[filterInd, filterPos]
  
  # Imputation of remaining NA values with major alleles.
  MajorAllele <- apply(xChar, 2,
                       function(x) names(sort(table(x), decreasing=TRUE))[1])
  MissingPos  <- which(is.na(xChar), arr.ind = TRUE)
  xImp  <- xChar
  xImp[MissingPos] <- MajorAllele[MissingPos[,2]]
  stopifnot(sum(is.na(xImp)) == 0)
  
  # Now that NA are removed, select variable sites.
  variablePos <- which(unlist(apply(xImp, 2, function(x) length(table(x)))) > 1)
  stopifnot(length(variablePos) > 0)
  
  # Just to keep track how many of the variable positions got imputed in at least on individual
  numImputed <- sum(variablePos %in% MissingPos[,2])
  x2 <- as.DNAbin(xImp[, variablePos, drop=FALSE])
  
  # Average number of differences between 2 sequences from different populations
  xPop <- factor(pop[labels(x2), 1])
  haploFreqs <- haploFreq(x2, xPop)
  numPop <- dim(haploFreqs)[2]
  rownames(haploFreqs) <- labels(haplotype(x2))
  haploDists <- as.matrix(dist.hamming(haplotype(x2)))
  popuDist <- data.frame(pop1 = factor(NA, levels = colnames(haploFreqs)),
                         pop2 = factor(NA, levels = colnames(haploFreqs)),
                         Dist = 0)
  i = 0
  for (popu1 in colnames(haploFreqs)) {
    for (popu2 in colnames(haploFreqs)) {
      if (popu1 != popu2) {
        i <- i + 1
        popuDist[i, 'pop1'] <- popu1
        popuDist[i, 'pop2'] <- popu2
        popuDist[i, 'Dist'] <- 0
        for (hap1 in labels(haplotype(x2))) {
          for (hap2 in labels(haplotype(x2))) {
            if (hap1 != hap2) {
               popuDist[i, 'Dist'] <- popuDist[i, 'Dist'] +
                 haploFreqs[hap1, popu1] * haploFreqs[hap2, popu2] *
                 haploDists[hap1, hap2]
            }
          }
        }
        popuDist[i, 'Dist'] <- popuDist[i, 'Dist'] *
          1 / (sum(haploFreqs[, popu1]) * sum(haploFreqs[, popu2]))
      }
    }
  }
  
  # Number of segregating sites per population
  popuSites <- data.frame(SegSites = numeric(length=numPop))
  rownames(popuSites) <- colnames(haploFreqs)
  for (popu in colnames(haploFreqs)) {
    filter <- pop[rownames(x2), 1] == popu
    popuSites[popu, 'SegSites'] <- sum(site.spectrum(x2[filter,]))
  }
  
  # The values to return are: number of variable sites, number of variable sites
  # per population (n), distances between pairs of populations n^2 - n (sic),
  # number of individuals per population, original alignment length (without gap
  # columns), variable positions with any imputation. That's n + 4 + n^2
  
  Final <- vector('list', length = numPop + 3 + numPop * numPop)
  names(Final)[1] <- 'BlockLength'
  Final[[1]] <- dim(xImp)[2]  # Length of alignment without gap columns.
  names(Final)[2] <- 'VariablePos'
  Final[[2]] <- length(variablePos)   # Number of variable positions overall.
  names(Final)[3] <- 'withImputation'
  Final[[3]] <- numImputed    # Number of variable positions with any imputation.
  for (i in 1:numPop) {
    names(Final)[i + 3] <- sprintf("%s_sites", rownames(popuSites)[i])
    Final[[i + 3]] <- popuSites[i, 'SegSites']
  }
  for (i in 1:numPop) {
    popu <- rownames(popuSites)[i]
    names(Final)[i + 3 + numPop] <- sprintf("%s_indiv", popu)
    Final[[i + 3 + numPop]] <- sum(pop[rownames(x2), 1] == popu)
  }
  i <- 3 + 2 * numPop
  for (pop1 in rownames(popuSites)) {
    for (pop2 in rownames(popuSites)) {
      if (pop1 != pop2) {
        i <- i + 1
        names(Final)[i] <- sprintf("Dist_%s_%s", pop1, pop2)
        Final[[i]] <- popuDist[popuDist$pop1 == pop1 & popuDist$pop2 == pop2, 'Dist']
      }
    }
  }
  return(Final)
}

mkaTable <- function(pop, fastaDir='fasta', fastaList=dir(fastaDir), .pb=NULL) {
  results <- data.frame()
  for (fastafile in fastaList) {
    update_progress(.pb)
    aln <- fasta2DNAbin(sprintf("%s/%s", fastaDir, fastafile), quiet = TRUE)
    try(z <- diverdiver(aln, pop), silent = TRUE)
    if ('z' %in% ls()) {
      if ((length(z) == ncol(results)) || (nrow(results) == 0)) {
        results <- rbind(results, z)
        rownames(results)[dim(results)[1]] <- substr(fastafile, 1, 7)
      }
      rm(z)
    }
    rm(aln)
  }
  return(results)
}
```

```{r loop}
if (! 'mka.RData' %in% dir()) {
  pb <- progress_estimated(length(dir('fasta')))
  pop3 <- pop2[pop2$pop %in% c('roumanicus','europaeus','concolor'), ,drop=FALSE]
  Results <- mkaTable(pop3, .pb=pb)
} else {
  load('mka.RData')
}

save(Results, file='mka.RData')
```




