---
title: "Untitled"
author: "J. Ignacio Lucas Lledó"
date: "12/2/2020"
output: html_document
params:
   MAPPING1: 'original_mapping_statistics.txt'
   MAPPING2: 'new_mapping_statistics.txt'
   VCF1: '/data/kristyna/hedgehog/results_2020/2020-02-02/russia.vcf'
   VCF2: '/data/kristyna/hedgehog/results_2020/2020-02-02/2020-02-02.recode.vcf'
---

```{r setup, message=FALSE}
library(ggplot2)
library(tidyr)
library(vcfR)
```

# Mapping results

```{r mapping1}
mapping1 <- read.table(params$MAPPING1, header=TRUE, sep='\t')
mapping1_long <- pivot_longer(mapping1[,-5], c(2,3,4), names_to='fate', values_to='reads')
mapping1_long$setting <- factor('very sensitive', levels=c('very sensitive', 'sensitive'))
```

```{r mapping2, fig.width=15}
mapping2 <- read.table(params$MAPPING2, header=TRUE, sep='\t')
mapping2_long <- pivot_longer(mapping2[,-5], c(2,3,4), names_to='fate', values_to='reads')
mapping2_long$setting <- factor('sensitive', levels=c('very_sensitive', 'sensitive'))

joined <- rbind(mapping1_long, mapping2_long)

ggplot(data=joined, mapping=aes(x=Sample, y=reads, fill=fate)) +
   geom_bar(stat='identity') +
   theme(axis.text.x = element_text(angle=90)) +
   facet_wrap(~setting)

ggplot(data=joined, mapping=aes(x=setting, y=reads, fill=fate)) +
   geom_bar(stat='identity')
```

The global proportion of reads mapped once is `r sum(mapping1$Mapped_once)/sum(mapping1$Total)`
in the very sensitive mapping and `r sum(mapping2$Mapped_once)/sum(mapping2$Total)` in the just
sensitive mapping. Not a big difference.

# Filtering of variant sites

The other concern is the filtering of variant sites, which I think is quite critical. The vcf
files contain information on global depth (across samples), which is important. It would be
a good chance to learn the usage of vcfR. But after trying a bit, I find it slow and difficult
to extract the depth information. I'd rather use `gawk` to get depth information, which is what
I need now. Now, how do I use a variable in the R namespace (`params$VCF1`) inside a `gawk` chunk?
Apparently, I can pass arguments to gawk through the chunk's options [https://github.com/yihui/knitr-examples/blob/master/024-engine-awk.Rmd]. But I could not make that
run. I resort to the `README.sh`, where I get the tables I need, and then use R to plot results,
as in the old times.

```{r depth}
TotalDepth <- read.table('total_depth.txt', col.names=c('DP','original', 'filtered'))
ggplot(data=TotalDepth, mapping=aes(x=DP)) +
   scale_x_log10() +
   geom_line(aes(y=original), color='blue') +
   geom_line(aes(y=filtered), color='red') +
   xlab('Depth') + ylab('frequency')
```

The shape of the distribution of total depths in the filtered vcf looks fine. The most remarkable
issue may be the magnitude of data loss during filtering. Vcftools does not allow to filter by
total depth, but only by individual sample's depth. The limits used were 4 and 200 per sample, and
a minimum of 75% of the 40 samples should be present. Thus, theoretical whole depth limits were
`r 40 * 0.75 * 4` and `r 40 * 200`. The actual minimum and maximum depth values observed in the
filtered vcf file are `r min(TotalDepth$DP[TotalDepth$filtered > 0])` and 
`r max(TotalDepth$DP[TotalDepth$filtered > 0])`.

It is possible that we are wasting data by the sample-specific maximum depth. I would trust most
sites where total depth is below 1000, even if some samples are over 200. I would like to quantify that.
For that, I need individual sample's depth data, now available in file `all_depths.txt`.

```{r depths}
depths <- read.table('all_depths.txt', header=TRUE)
# Sites with at least 30 samples with depths between 4 and 200:
min4max200 <- rowSums(depths[,2:41] >= 4 & depths[,2:41] <= 200) >= 30
plot(table(depths[,1]), log='x', type='l', col='blue', xlab='Depth', ylab='frequency')
points(table(depths[min4max200,1]), col='red', type='l')
```

The maximum depth allowed per sample of 200 is doing a good job at keeping most
sites with a reasonable coverage, without admitting too many of the excessively
covered ones. We loose sites on the low coverage end, which could only be helped
by lowering the minimum coverage allowed per sample, or relaxing the percentage
of samples required to be covered (75% currently). None of these options seem
a good idea.

In the dataset from central European contact zone, population assignment was
clear, because very few hybrids were present. There, we were able to use species-
specific quotas (e.g., at least 10 *E. europaeus* and 10 *E. roumanicus* samples
covered per site), which optimizes the selection of sites. But we cannot do the
same when hybridization seems much more frequent.

The difference in height of the filtered sites (in red) between the two plots
indicates we loose many sites from analysis for reasons other than the coverage.
The filtering step requires also a minimum allele frequency of 0.01, a maximum
allele frequency of 0.95, neither more nor less than 2 alleles, a minimum quality
of 50 and a spacing of at least 261 nucleotides between SNPs. I would not change
any of those parameters, but perhaps the maximum allele frequency. With 40 samples,
the precision in allele frequency is `r 1/80`. I think allele frequency limits
should be symmetrical, like 0.0125 and 0.9875, in order to admit singleton alleles
of both types, reference and alternative.

```{r sessionInfo}
save(joined, TotalDepth, depths, file='assessment.RData')
sessionInfo()
```
