---
title: "Diversity and divergence in E. europaeus, E. roumanicus and E. concolor"
author: "J. Ignacio Lucas Lledó"
date: "7/4/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(gridExtra)
```

# The data
I read the data in three main data frames:

- `allStats` includes the genome-wide population genetics for the three species and the
introgression statistics for the pair *E. roumanicus* and *E. europaeus*.
- `geneStats` has the population genetics statistics for the three species
only in genic regions: diversity ($\pi$), divergence ($d_{xy}$), and fixation
index ($F_{ST}$).
- `interStats` has the population genetics statistics for the three species
only in intergenic regions.

```{r data}
popgen     <- read.table('nonoverlap.PopGenStats.csv',  header=TRUE, sep=',',
                         col.names=c('scaffold', 'start', 'end', 'mid', 'sites',
                                     'pi_rou', 'pi_eur', 'pi_con',
                                     'dxy_rou_eur', 'dxy_rou_con', 'dxy_eur_con',
                                     'Fst_rou_eur', 'Fst_rou_con', 'Fst_eur_con'))
abbababa   <- read.table('nonoverlap.abbababa.csv', header=TRUE, sep=',')
abbababa$fd[abbababa$D <= 0] <- 0.0
abbababa$fdM[abbababa$D <= 0] <- 0.0
allStats   <- merge(popgen, abbababa, by=c('scaffold', 'start', 'end'))
allStats   <- allStats[! is.na(allStats$fd),]
geneStats  <- read.table('genes.PopGenStats.csv', header=TRUE, sep=',',
                         col.names=c('scaffold', 'start', 'end', 'mid', 'sites',
                                     'pi_rou', 'pi_eur', 'pi_con',
                                     'dxy_rou_eur', 'dxy_rou_con', 'dxy_eur_con',
                                     'Fst_rou_eur', 'Fst_rou_con', 'Fst_eur_con'))
interStats <- read.table('inter.PopGenStats.csv', header=TRUE, sep=',',
                         col.names=c('scaffold', 'start', 'end', 'mid', 'sites',
                                     'pi_rou', 'pi_eur', 'pi_con',
                                     'dxy_rou_eur', 'dxy_rou_con', 'dxy_eur_con',
                                     'Fst_rou_eur', 'Fst_rou_con', 'Fst_eur_con'))
```

# *E. roumanicus* and *E. europaeus*

First I model the divergence between *E. roumanicus* and *E. europaeus* as a
set of linear relationships with the rellevant genetic diversities and with
the measure of genetic introgression $f_d$. Note that no introgression is
detected in about 38% of sites, where $f_d = 0$.

```{r linearModels}
m1 <- lm(dxy_rou_eur ~ pi_eur,               data=allStats)
m2 <- lm(dxy_rou_eur ~ pi_rou,               data=allStats)
m3 <- lm(dxy_rou_eur ~ pi_eur + fd,          data=allStats)
m4 <- lm(dxy_rou_eur ~ pi_rou + pi_eur,      data=allStats)
m5 <- lm(dxy_rou_eur ~ pi_rou + fd,          data=allStats)
m6 <- lm(dxy_rou_eur ~ pi_rou + pi_eur + fd, data=allStats)

summary(m6)

allStats$res3 <- residuals(m3)
allStats$res5 <- residuals(m5)
```

Previous observation that the divergence between *E. roumanicus* and *E. europaeus*
shows a negative correlation with the genetic diversity within *E. roumanicus* is
maintained. We can also see that the relationship with the diversity in *E. europaeus*
is also negative, although much less important.

Using the residuals from model 3, we remove the effect of the detected introgression
(*fd*), and the effect of diversity in *E. europaeus*. As we can see below, even after
removing those effects, a considerable fraction of the variation in divergence among
genomic sites could be explained by levels of genetic variation in *E. roumanicus*,
and the relationship is negative.

```{r plotsRouEur, message=FALSE, fig.width=10}
p1 <- ggplot(allStats, aes(x=pi_rou, y=dxy_rou_eur, color=fd)) +
  geom_point() + geom_smooth(method='lm') + xlab('Diversity in E. roumanicus') +
  ylab('Divergence') + scale_colour_gradient2(low='yellow',mid='orange',high='red',midpoint=0.1) +
  ggtitle('Including the effect of fd')
p2 <- ggplot(allStats, aes(x=pi_rou, y=res3)) +
  geom_point() + geom_smooth(method='lm') + xlab('Diversity in E. roumanicus') +
  ylab('Divergence residuals (m3)') + ggtitle('Removing the effect of fd')
grid.arrange(p1, p2, nrow=1)
```

The fact that our measure of introgression does not remove completely the
negative correlation suggests two interpretations: either our measure
of introgression badly underestimates introgression, or there is something
else going one. Both options are compatible.

```{r GenesRouEur, message=FALSE, fig.width=10}
p1 <- ggplot(geneStats, aes(x=pi_rou, y=dxy_rou_eur)) +
  geom_point() + geom_smooth(method='lm') + xlab('Diversity in E. roumanicus') +
  ylab('Divergence with E. europaeus') + ggtitle('Genic regions')
p2 <- ggplot(interStats, aes(x=pi_rou, y=dxy_rou_eur)) +
  geom_point() + geom_smooth(method='lm') + xlab('Diversity in E. roumanicus') +
  ylab('Divergence with E. europaeus') + ggtitle('Intergenic regions')
grid.arrange(p1, p2, nrow=1)
```

The fact that the above negative correlation is stronger in intergenic regions
could point at a functional effect. But could also be a consequence of the larger
number of data points. I would not give it much importance.

What about the correlation between diversity in *E. europaeus* and its divergence
with *E. roumanicus*? The linear models suggested an also negative, but milder
relationship.

```{r withEuropaeus, message=FALSE, fig.width=10, fig.height=10}
p1 <- ggplot(allStats, aes(x=pi_eur, y=dxy_rou_eur, color=fd)) + geom_point() +
  geom_smooth(method='lm') + xlab('Diversity in E. europaeus') +
  scale_colour_gradient2(low='yellow', mid='orange', high='red',midpoint=0.1) +
  ylab('Divergence with E. roumanicus') + ggtitle('Including the effect of fd')
p2 <- ggplot(allStats, aes(x=pi_eur, y=res5)) + geom_point() +
  geom_smooth(method='lm') + xlab('Diversity in E. europaeus') +
  ylab('Divergence with E. roumanicus') + ggtitle('Excluding the effect of fd')
p3 <- ggplot(geneStats, aes(x=pi_eur, y=dxy_rou_eur)) + geom_point() +
  geom_smooth(method='lm') + xlab('Diversity in E. europaeus') +
  ylab('Divergence with E. roumanicus') + ggtitle('Genic regions')
p4 <- ggplot(interStats, aes(x=pi_eur, y=dxy_rou_eur)) + geom_point() +
  geom_smooth(method='lm') + xlab('Diversity in E. europaeus') +
  ylab('Divergence with E. roumanicus') + ggtitle('Intergenic regions')
grid.arrange(p1, p2, p3, p4, nrow=2)
```

The lack of correlation is also a departure from the neutral expectations, and
requires an explanation. I guess, we could claim introgression as well.

# *E. roumanicus* and *E. concolor*

We can do the same for *E. roumanicus* and *E. concolor*. I include the estimated
introgression between *E. roumanicus* and *E. europaeus* in the model, because it
could affect positively the divergence between *E. roumanicus* and *E. concolor*.

```{r RoumanicusConcolor}
m1 <- lm(dxy_rou_con ~ pi_rou,               data=allStats)
m2 <- lm(dxy_rou_con ~ pi_con,               data=allStats)
m3 <- lm(dxy_rou_con ~ pi_rou + pi_con,      data=allStats)
m4 <- lm(dxy_rou_con ~ pi_rou + fd,          data=allStats)
m5 <- lm(dxy_rou_con ~ pi_con + fd,          data=allStats)
m6 <- lm(dxy_rou_con ~ pi_rou + pi_con + fd, data=allStats)

summary(m6)
```

All terms in `m6` are significant and with a positive effect:

```{r plotsRouCon, message=FALSE, fig.width=10}
p1 <- ggplot(allStats, aes(x=pi_rou, y=dxy_rou_con)) +
  geom_point() + geom_smooth(method='lm') + xlab('Diversity in E. roumanicus') +
  ylab('Divergence with E. concolor')
p2 <- ggplot(allStats, aes(x=pi_con, y=dxy_rou_con)) +
  geom_point() + geom_smooth(method='lm') + xlab('Diversity in E. concolor') +
  ylab('Divergence with E. roumanicus')
grid.arrange(p1, p2, nrow=1)
```

The positive relationship is exactly what we expect from the neutral theory.

```{r GenesRouCon, message=FALSE, fig.width=10}
p1 <- ggplot(geneStats, aes(x=pi_rou, y=dxy_rou_con)) + geom_point() +
  geom_smooth(method='lm') + xlab('Diversity in E. roumanicus') +
  ylab('Divergence wih E. concolor') + ggtitle('Genic regions')
p2 <- ggplot(interStats, aes(x=pi_rou, y=dxy_rou_con)) + geom_point() +
  geom_smooth(method='lm') + xlab('Diversity in E. roumanicus') +
  ylab('Divergence with E. concolor') + ggtitle('Intergenic regions')
grid.arrange(p1, p2, nrow=1)
```

According to this, the divesity seggregating in *E. roumanicus*, even in
genic regions, is quite neutral. Recall that we actually sampled very few
non-synonymous polymorphisms, and we did not distinguish exons from introns
within genic regions. Our functional classification of sites could be too
rough to notice the selective constraints.

# *E. europaeus* and *E. concolor*

The analysis would not be complete without the last comparison. Because there is
clearly, absolutely no introgression between *E. europaeus* and *E. concolor*, the
expected pattern is the same as between *E. roumanicus* and *E. concolor*: a positive
relationship between diversity and divergence. I skip the linear models.

```{r EurCon, message=FALSE, fig.width=10, fig.height=10}
p1 <- ggplot(allStats, aes(x=pi_eur, y=dxy_eur_con, color=fd)) + geom_point() +
  geom_smooth(method='lm') + xlab('Diversity in E. europaeus') +
  scale_colour_gradient2(low='yellow', mid='orange', high='red',midpoint=0.1) +
  ylab('Divergence with E. concolor')
p2 <- ggplot(allStats, aes(x=pi_con, y=dxy_eur_con, color=fd)) + geom_point() +
  geom_smooth(method='lm') + xlab('Diversity in E. concolor') +
  scale_color_gradient2(low='yellow', mid='orange', high='red', midpoint=0.1) +
  ylab('Divergence with E. europaeus')
p3 <- ggplot(geneStats, aes(x=pi_eur, y=dxy_eur_con)) + geom_point() +
  geom_smooth(method='lm') + xlab('Diversity in E. europaeus') +
  ylab('Divergence with E. concolor') + ggtitle('Genic regions')
p4 <- ggplot(interStats, aes(x=pi_eur, y=dxy_eur_con)) + geom_point() +
  geom_smooth(method='lm') + xlab('Diversity in E. europaeus') +
  ylab('Divergence with E. concolor') + ggtitle('Intergenic regions')
grid.arrange(p1, p2, p3, p4, nrow=2)
```

This unexpected lack of correlation makes me think that introgression is
not the end of the story. I do not see a way to explain this with introgression.
There must be something else going on here. Now, it looks like *E. europaeus* is
peculiar. Its main difference may be its population structure. I have not been
able yet to imagine a way for the population structure to affect the relationship
between divergence and diversity.

# Some more plots

```{r boxplots}
boxplot(list(E.concolor   = allStats$pi_con,
             E.roumanicus = allStats$pi_rou,
             E.europaeus  = allStats$pi_eur),
        ylab = 'Genetic diversity',
        notch = TRUE)

boxplot(list(roumanicus_concolor  = allStats$dxy_rou_con,
             roumanicus_europaeus = allStats$dxy_rou_eur,
             europaeus_concolor   = allStats$dxy_eur_con),
        ylab = 'Between species divergence',
        notch = TRUE)
```

# Session Information
```{r sessionInfo}
sessionInfo()
```
