---
title: "Asymptotic MK"
author: "J. Ignacio Lucas Lledó"
date: "26/2/2020"
output: html_document
bibliography:
  asymptoticMK.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

I downloaded the R functions required to run the asymptotic MK test localy from
[here](https://raw.githubusercontent.com/MesserLab/asymptoticMK/master/asymptoticMK_local.R),
as suggested by [@Messer2013]. I had to edit one line near the bottom, to set a directory
where to save some plots. I also comment out the demo execution of `asymptoticMK.Drosophila()`,
and prevent the change of working directory.

```{r source, results='hide'}
library(ggplot2)
source('../../bin/asymptoticMK_local.R')
```

```{r alphas}
combineRows <- function(df){
  # Takes a polymorphisms data frame, and combines rows with zero
  # values with the following row (or the previous, if last), and
  # assigns the resulting number of polymorphims to the average
  # frequency class. It's not perfect, but may suffice.
  while (any(rowSums(df[,2:3] == 0) > 0)) {
    zeroIndex <- which(rowSums(df[,2:3] == 0) > 0)[1]
    if (zeroIndex < dim(df)[1]) {
      df[zeroIndex + 1, 1] <- mean(c(df[zeroIndex,1], df[zeroIndex + 1,1]))
      df[zeroIndex + 1, 2:3] <- df[zeroIndex,2:3] + df[zeroIndex + 1,2:3]
    } else {
      df[zeroIndex - 1, 1] <- mean(c(df[zeroIndex,1], df[zeroIndex - 1,1]))
      df[zeroIndex - 1, 2:3] <- df[zeroIndex,2:3] + df[zeroIndex - 1,2:3]
    }
    df <- df[-zeroIndex,]
    row.names(df) <- 1:dim(df)[1]
  }
  return(df)
}
minSamples <- 7:12
EuropaeusMK <- data.frame()
RoumanicusMK <- data.frame()
for (i in minSamples) {
  z <- as.matrix(read.table(sprintf('E%gR1.divergence.txt', i),
                            colClasses = c('NULL','NULL','NULL','numeric')))
  EuropaeusSynonymousDivergence <- z[1]
  EuropaeusMissenseDivergence   <- z[2]
  z <- as.matrix(read.table(sprintf('R%gE1.divergence.txt', i),
                            colClasses = c('NULL','NULL','NULL','numeric')))
  RoumanicusSynonymousDivergence <- z[1]
  RoumanicusMissenseDivergence   <- z[2]
  EuroPolymorphism <- read.table(sprintf('E%gR1.polymorphism.txt',i), col.names=c('f','p','p0'))
  EuroPolymorphism <- combineRows(EuroPolymorphism)   # combines rows with zero values with next or previous ones.
  RoumaPolymorphism <- read.table(sprintf('R%gE1.polymorphism.txt',i), col.names=c('f','p','p0'))
  RoumaPolymorphism <- combineRows(RoumaPolymorphism)
  EuropaeusMK <- rbind(EuropaeusMK, asymptoticMK(EuropaeusSynonymousDivergence,
                                                 EuropaeusMissenseDivergence,
                                                 0.1, 0.9, EuroPolymorphism))
  RoumanicusMK <- rbind(RoumanicusMK, asymptoticMK(RoumanicusSynonymousDivergence,
                                                   RoumanicusMissenseDivergence,
                                                   0.1, 0.9, RoumaPolymorphism))
}
EuropaeusMK$MinSamples <- minSamples
RoumanicusMK$MinSamples <- minSamples
EuropaeusMK$Species <- 'E. europaeus'
RoumanicusMK$Species <- 'E. roumanicus'
AMK <- rbind(EuropaeusMK, RoumanicusMK)
```

```{r plot}
ggplot(data=AMK, mapping=aes(x=MinSamples, y=alpha_asymptotic)) +
  geom_ribbon(aes(ymin=CI_low, ymax=CI_high), fill=3, alpha=0.5) +
  geom_line() + facet_wrap(~Species)
```

Even using all possible sites with a minimum number of samples of the focal species
genotyped, and requiring at least 1 genotyped individual from the sister species
(to estimate divergence), the numbers of sites are low, and the asymptotic estimates
of $\alpha$, very noisy. Of course, 95% confidence intervals beyond (-1,1) are
unnecessarily wide. 

# References