---
title: "ANGSD"
author: "J. Ignacio Lucas Lledó"
date: "6/12/2021"
output: html_document
bibliography: references.bib
---

## Becoming familiar with ANGSD

ANGSD [@Nielsen2012] can perform several analysis using genotype likelihoods, including the
ABBA/BABA statistics. Because it does not require calling genotypes, the input
data can be sequence data in BAM files. Alternatively, it accepts genotype
likelihoods in Beagle format, and even VCF. But the ABBA/BABA test and others
can only be performed from sequence files. I do not want to pass row BAM files
to `angsd`, because it is unclear how it handles indels, which are abundant
in our data set. I trust `freebayes` to align indels better, and that's why
I prefer to use the genotype likelihoods for only binary SNPs stored in our
VCFs.

PCAngsd is a separate program in Python. Unfortunately, I was not able to
install it system-wide, and I'm using below my local installation.

## Estimate allele frequencies and site-frequency spectra.

Allele frequencies can be estimated from VCF, for every site. But I don't
see the way to filter by individuals. I must create a population-specific
vcf, from the original `../2021-11-16/all.vcf`, so that variable sites are
a representative sample of variation in each species. Recall that the filtered
VCF in `2021-12-01` required sites to be variable overall, which introduces
many fixed differences among species that would bias their SFS.

ANGSD does not accept just any VCF. From what I experienced it occasionally
stops at an intermediate point of a VCF when estimating allele frequencies or
site-frequency spectra. This is a sign of some sites not having the expected
quality or amount of data. I solved the issue by requiring genotypes to be
covered at least 5 times (`--minDP 5`; maybe 4 would have been enough) and
sites to have at least 1 minor allele counted (`--mac 1`).

```{bash splitVCF}
if [ ! -e popfile.txt ]; then ./make_popfile.sh; fi
if [ ! -e lowest15.txt ]; then cp ../2021-12-01/lowest15.txt ./; fi

VCF='../2021-11-16/all.vcf'
for pop in roumanicus europaeus concolor; do
   if [ ! -e $pop.txt ]; then
      grep $pop popfile.txt | grep -v -f lowest15.txt | cut -f 1 > $pop.txt
   fi
   if [ ! -e $pop.recode.vcf ]; then
      vcftools --vcf $VCF \
               --keep $pop.txt \
               --remove-indels \
               --minQ 50 \
               --minDP 5 \
               --min-alleles 2 \
               --max-alleles 2 \
               --max-missing 0.75 \
               --mac 1 \
               --recode \
               --out $pop &> $pop.recode.log &
   fi
done
wait

for pop in roumanicus europaeus concolor; do
   if [ ! -e $pop.mafs ]; then
      angsd -vcf-gl $pop.recode.vcf \
            -fai ../../data/reference.fa.fai \
            -doMaf 1 -doMajorMinor 1 -out $pop &> $pop.mafs.log
      gunzip $pop.mafs.gz
   fi
   if [ ! -e $pop.sfs ]; then
      if [ ! -e $pop.saf.idx ]; then
         angsd -vcf-gl $pop.recode.vcf -doMajorMinor 1 -out $pop \
               -doSaf 1 -anc ../../data/reference.fa &> $pop.saf.idx.log
      fi
      ~/bin/angsd/misc/realSFS $pop.saf.idx \
            -maxIter 400 -P 4 -tole .00001 -fold 1 1> $pop.sfs 2> $pop.sfs.log
   fi
done
```

## VCF-derived site-frequency spectra

```{r plots}
concolor.sfs <- scan('concolor.sfs')
europaeus.sfs <- scan('europaeus.sfs')
roumanicus.sfs <- scan('roumanicus.sfs')
barplot(concolor.sfs[1:((length(concolor.sfs) - 1) / 2)],
        main = 'E. concolor', xlab = 'Minor allele count',
        ylab = 'Number of SNPs')
barplot(europaeus.sfs[1:((length(europaeus.sfs) - 1) / 2)],
        main = 'E. europaeus', xlab = 'Minor allele count',
        ylab = 'Number of SNPs')
barplot(roumanicus.sfs[1:((length(roumanicus.sfs) - 1)/ 2)],
        main = 'E. roumanicus', xlab = 'Minor allele count',
        ylab = 'Number of SNPs')
```

Sites had been required to be bi-allelic. But the class of monomorphic sites is not
empty in any of the SFS. The reason is just that even if sites look bi-allelic to
`vcftools` some of them have a probability of actually being monomorphic, which adds
up to several expected monomorphic sites in the whole data set. Actually,
requiring sites to be bi-allelic must be causing the lack of very low frequency
sites. This alone could be enough to explain the shape of the SFS in *E. roumanicus*.

In the case of *E. europaeus*, the SFS is intriguing. The second peak is a huge excess
of intermediate frequency SNPs. I suppose it could be caused by the presence of the
two lineages. 

The SFS in *E. concolor* is a complete mistery. I cannot make sense of it. I am manually
running `angsd -doSaf` with different parameters, and starting from BAM files, just to
see if the result is robust. The bootstrap already confirmed the shape.

```{r mafs}
library(ggplot2)
concolor.mafs   <- read.table('concolor.mafs',   header = TRUE, as.is = TRUE)
europaeus.mafs  <- read.table('europaeus.mafs',  header = TRUE, as.is = TRUE)
roumanicus.mafs <- read.table('roumanicus.mafs', header = TRUE, as.is = TRUE)
row.names(concolor.mafs)   <- paste(concolor.mafs$chromo,
                                    concolor.mafs$position, sep = '_')
row.names(europaeus.mafs)  <- paste(europaeus.mafs$chromo,
                                    europaeus.mafs$position, sep = '_')
row.names(roumanicus.mafs) <- paste(roumanicus.mafs$chromo,
                                    roumanicus.mafs$position, sep = '_')
concolor.mafs$first  <- pmin(concolor.mafs$major, concolor.mafs$minor)  # alphabetically first
concolor.mafs$second <- pmax(concolor.mafs$major, concolor.mafs$minor)  # alphabetically second
stopifnot(max(concolor.mafs$knownEM) <= 0.5)  # knownEM is the minor allele freq.
concolor.mafs$firstFreq <- sapply(1:dim(concolor.mafs)[1], function(x) {
  ifelse(concolor.mafs[x, 'minor'] < concolor.mafs[x, 'major'],
         concolor.mafs[x, 'knownEM'], 1.0 - concolor.mafs[x, 'knownEM'])
})

europaeus.mafs$first  <- pmin(europaeus.mafs$major, europaeus.mafs$minor)
europaeus.mafs$second <- pmax(europaeus.mafs$major, europaeus.mafs$minor)
stopifnot(max(europaeus.mafs$knownEM) <= 0.5)
europaeus.mafs$firstFreq <- sapply(1:dim(europaeus.mafs)[1], function(x) {
  ifelse(europaeus.mafs[x, 'minor'] < europaeus.mafs[x, 'major'],
         europaeus.mafs[x, 'knownEM'], 1.0 - europaeus.mafs[x, 'knownEM'])
})

roumanicus.mafs$first  <- pmin(roumanicus.mafs$major, roumanicus.mafs$minor)
roumanicus.mafs$second <- pmax(roumanicus.mafs$major, roumanicus.mafs$minor)
stopifnot(max(roumanicus.mafs$knownEM) <= 0.5)
roumanicus.mafs$firstFreq <- sapply(1:dim(roumanicus.mafs)[1], function(x) {
  ifelse(roumanicus.mafs[x, 'minor'] < roumanicus.mafs[x, 'major'],
         roumanicus.mafs[x, 'knownEM'], 1.0 - roumanicus.mafs[x, 'knownEM'])
})
common.ce <- intersect(row.names(concolor.mafs),
                       row.names(europaeus.mafs))
ce.mafs <- data.frame(
  c.first     = concolor.mafs[common.ce, 'first'],
  c.second    = concolor.mafs[common.ce, 'second'],
  c.firstFreq = concolor.mafs[common.ce, 'firstFreq'],
  e.first     = europaeus.mafs[common.ce, 'first'],
  e.second    = europaeus.mafs[common.ce, 'second'],
  e.firstFreq = europaeus.mafs[common.ce, 'firstFreq'],
  sameAlleles = concolor.mafs[common.ce, 'first'] == europaeus.mafs[common.ce, 'first'] &
                concolor.mafs[common.ce, 'second'] == europaeus.mafs[common.ce, 'second']
)

common.er <- intersect(row.names(europaeus.mafs),
                       row.names(roumanicus.mafs))
er.mafs <- data.frame(
  e.first     = europaeus.mafs[common.er, 'first'],
  e.second    = europaeus.mafs[common.er, 'second'],
  e.firstFreq = europaeus.mafs[common.er, 'firstFreq'],
  r.first     = roumanicus.mafs[common.er, 'first'],
  r.second    = roumanicus.mafs[common.er, 'second'],
  r.firstFreq = roumanicus.mafs[common.er, 'firstFreq'],
  sameAlleles = europaeus.mafs[common.er, 'first']  == roumanicus.mafs[common.er, 'first'] &&
                europaeus.mafs[common.er, 'second'] == roumanicus.mafs[common.er, 'second']
)

common.cr <- intersect(row.names(concolor.mafs),
                       row.names(roumanicus.mafs))
cr.mafs <- data.frame(
  c.first     = concolor.mafs[common.cr, 'first'],
  c.second    = concolor.mafs[common.cr, 'second'],
  c.firstFreq = concolor.mafs[common.cr, 'firstFreq'],
  r.first     = roumanicus.mafs[common.cr, 'first'],
  r.second    = roumanicus.mafs[common.cr, 'second'],
  r.firstFreq = roumanicus.mafs[common.cr, 'firstFreq'],
  sameAlleles = concolor.mafs[common.cr, 'first']  == roumanicus.mafs[common.cr, 'first'] &&
                concolor.mafs[common.cr, 'second'] == roumanicus.mafs[common.cr, 'second']
)

ggplot(data = ce.mafs[ce.mafs$sameAlleles, ], mapping = aes(x = c.firstFreq, y = e.firstFreq)) +
  geom_hex() + scale_fill_gradient(trans = 'log') +
  xlab('Freq. in E. concolor') + ylab('Freq in E. europaeus')

ggplot(data = er.mafs[er.mafs$sameAlleles, ], mapping = aes(x = r.firstFreq, y = e.firstFreq)) +
  geom_hex() + scale_fill_gradient(trans = 'log') +
  xlab('Freq. in E. roumanicus') + ylab('Freq. in E. europaeus')

ggplot(data = cr.mafs[cr.mafs$sameAlleles, ], mapping = aes(x = c.firstFreq, y = r.firstFreq)) +
  geom_hex() + scale_fill_gradient(trans = 'log') +
  xlab('Freq. in E. concolor') + ylab('Freq. in E. roumanicus')

```

One difficulty to interprete the plots above is that they compare the frequency of
the alphabetically first allele between species. That was easier than guessing the
ancestral one. But, I will probably have to infer the ancestral alleles and unfold
the SFS. Next folder.

```{r densities}
plot(density(roumanicus.mafs$knownEM), col = 'blue', lwd = 2, xlab = 'Minor allele freq.')
lines(density(europaeus.mafs$knownEM), col = 'red',  lwd = 2)
lines(density(concolor.mafs$knownEM), col = 'green', lwd = 2)
``` 

The smoothed densities of minor allele frequencies resemble the respective SFSs, but
are not accurate. Just for comparison.

## Session Info

```{r sessioninfo}
sessionInfo()
```

## References